Signatures and Induction Principles for Higher Inductive-Inductive Types

### 2019-02-04 paper.tex
small change
stuff

The present paper is an expanded version of our conference paper \cite{hiit}.
In this version, we additionally compute notions of homomorphisms from
signatures, and compare them to logical relations, in Section
\ref{sec:morphisms}. We also explain a coherence problem in interpreting
syntaxes of type theories, and how this influenced the current paper, in Section
\ref{sec:coherence} and Section \ref{sec:categorical}.

@@

Cartmell's generalized algebraic theories (GATs) \cite{gat} pioneered a
type-theoretic notion of algebraic signature. GATs can be viewed as finitary
quotient inductive-inductive signatures (QIITs), although GATs support equalities
between inductive types (sorts) as well, which have not been so far explored in
other works about QIITs and HIITs.

@@

@@

(5) We have a function space where the domain is a type in the external
theory. We distinguish it from (3) by using red brick color in the domain
specification. We specify lists and the integers as follows.
\begin{alignat*}{4}
  & \hat{A}\in\Type_{\hat{0}} && \vdash\,\boldsymbol{\cdot},\,\, && List:\U,\,\,\,nil:\underline{List},\,\,\,cons:(\hat{x}\in \hat{A})\ra List\ra\underline{List} \\
  & \hat{\Gamma} && \vdash\,\boldsymbol{\cdot},\,\,&& Int:\U,\,\,\,pair:(\hat{x}\,\hat{y}\in\hat{Nat})\ra\underline{Int},\,\,\, \\
  & && && eq:(\hat{a}\,\hat{b}\,\hat{c}\,\hat{d}\in \hat{Nat})(\hat{p}\in \hat{a}\hat{\mathbin{+}}\hat{d}\eqh_{\hat{Nat}} \hat{b}\hat{\mathbin{+}}\hat{c})\ra \underline{pair\,\hat{a}\,\hat{b}=_{Int}\mathsf{pair}\,\hat{c}\,\hat{d}}, \\
  & && && trunc:(x y : Int)(p\,q : a=_{Int} b)\ra \underline{p=_{x=_{Int} y} q}
\end{alignat*}
In the case of integers, $\hat{\Gamma}$ is
$\hat{Nat}\in\Type_{\hat{0}}\hat{,}\,\blank\hat{\mathbin{+}}\blank\in\hat{Nat}\rah
\hat{Nat}\rah\hat{Nat}$, or alternatively, we could require natural
numbers in the external theory. As another example, propositional
truncation for a type $\hat{A}$ is specified as follows.
\[
\hat{A}\in\Type_{\hat{0}}\vdash\, \boldsymbol{\cdot},\,\,\,tr:\U,\,\,\,emb : (\hat{x}\in \hat{A})\ra \underline{tr},\,\,\,eq:(x\,y:tr)\ra \underline{x=_{tr} y}
\]
The smallness of $\hat{A}$ is required in (5). It is possible to generalize
signatures to arbitrary universe levels, but it is not essential to the current
development. Note that the (5) function space preserves strict positivity, since
in the external theory there is no way to recursively refer to the inductive type
\emph{being defined}. The situation is analogous to the case of $W$-types
\cite{abbot05containers}, where shapes and positions can contain arbitrary types
but they cannot recursively refer to the $W$-type being defined.

(6) $\U$ is also closed under a function space where the domain is an external
type and the codomain is a small source theory type. We overload the application
notation for external parameters, as it is usually clear from context which
application is meant. The rules allow types with infinitary constructors, for
example trees containing elements of $\hat{A}$ at the leaves and branching by
$\hat{B}$ (which could be an infinite type):
\[
\hat{A}\in\Type_{\hat{0}}\commah\hat{B}\in\Type_{\hat{0}}\vdash\,\boldsymbol{\cdot},\,\,\,T:\U,\,\,\,leaf:(\hat{x}\in \hat{A})\ra\underline{T},\,\,\,node:((\hat{x}\in \hat{B})\ra T)\ra\underline{T}
\]
Here, $leaf$ has a function type (5) and $node$ has a function type (3) with a
function type (6) in the domain. More generally, we can define $W$-types
\cite{abbot05containers} as follows. $\hat{S}$ describes the ``shapes'' of the
constructors and $\hat{P}$ the ``positions'' where recursive arguments can
appear.
\[
\hat{S}\in\Type_{\hat{0}}\commah\hat{P}:\hat{S}\rah\Type_{\hat{0}}\vdash\,\boldsymbol{\cdot},\,\,\,W:\U,\,\,\,sup: (\hat{s} \in \hat{S})\ra((\hat{p} \in \hat{P}\,\hat{s})\ra W)\ra \underline{W}
\]
For a more complex infinitary example, see the definition of Cauchy
reals in \cite[Definition 11.3.2]{HoTTbook}. It can be also found as
an example file in our Haskell implementation.

@@

\begingroup
\allowdisplaybreaks
\begin{alignat*}{5}
  & \boldsymbol{\cdot}^\S\,\hat{\gamma}\,\hat{\gamma^D} && :\equiv \hat{\top} \\
  & (\Delta,\,x:A)^\S\,(\hat{\gamma},\hat{t})\,(\hat{\gamma^D},\hat{t^D}) && :\equiv \hat{(}\hat{\gamma^S}\in\Delta^\S\hat{\gamma^2}\hat{)}\timesh A^\S\,\hat{\gamma^S}\,\hat{t}\,\hat{t^D} \\
  & x^\S\,\hat{\gamma^S} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma^S} \\
  & \U^\S\,\hat{\gamma^S}\,\hat{A}\,\hat{A^D} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  \hat{A^D}\,\hat{x} \\
  & (\underline{a})^\S\,\hat{\gamma^S}\,\hat{t}\,\hat{t^D} && :\equiv a^\S\,\hat{\gamma^S}\,\hat{t} \eqh \hat{t^D} \\
  & ((x:a)\ra B)^\S\,\hat{\gamma^S}\,\hat{f}\,\hat{f^D} && :\equiv \hat{(}\hat{x}\in a^\A\,\hat{\gamma}\hat{)}\rah  B^\S\,\hat{(}\hat{\gamma,x}\hat{)}\,\hat{(}\hat{\gamma^D,}\,a^\S\,\hat{\gamma^S}\,\hat{x}\hat{)}\,\hat{(}\hat{\gamma^S,}\,\hat{\refl}\hat{)} \\
  & && \hspace{8.4em} \hat{(}\hat{f}\,\hat{x}\hat{)}\,\hat{\big(}\hat{f^D}\,\hat{x}\,\hat{(}a^\S\,\hat{\gamma^S}\,\hat{x}\hat{)}\hat{\big)} \\
  & (t\,u)^\S\,\hat{\gamma^S} && :\equiv \hat{\J}\,\hat{(}t^\S\,\hat{\gamma^S}\,\hat{(} u^\A\,\hat{\gamma}\hat{)}\hat{)}\,\hat{(}u^\S\,\hat{\gamma^S}\hat{)} \\
  & ((\hat{x}\in \hat{A})\ra B)^\S\,\hat{\gamma^S}\,\hat{f}\,\hat{f^D} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah B^\S\,\hat{\gamma^S}\,\hat{(}\hat{f}\,\hat{x}\hat{)}\,\hat{(}\hat{f^D}\,\hat{x}\hat{)} \\
  & (t\,\hat{u})^\S\,\hat{\gamma^S} && :\equiv t^\S\,\hat{\gamma^S}\,\hat{u} \\
  & (t=_a u)^\S\,\hat{\gamma^S}\,\hat{e} && :\equiv \TR\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)}\hat{\big(}\TR\,\hat{(}u^\S\,\hat{\gamma^S}\hat{)}\,\hat{(}\apd\,\hat{(}a^\S\,\hat{\gamma^S}\hat{)}\,\hat{e}\hat{)}\hat{\big)} \\
  & (\refl_t)^\S\,\hat{\gamma^S} && :\equiv \hat{\J}\,\hat{\refl}\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)} \\
  & (\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\S\,\hat{\gamma^S} && :\equiv \\
  & && \hspace{-10em}\hat{\J}\,\hat{\bigg(}\hat{\J}\,\hat{\Big(}\hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}\hat{\lambda}\,\hat{p^D}\,\hat{p^S}\,\hat{pr^D}\,\hat{pr^S}\hat{.}\,\hat{pr^S}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)}\,\\
  & && \hspace{-9em}\hat{(}\hat{\mathsf{uncurry}}\,p^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\S\,\hat{\gamma^S}\hat{)}\,\hat{(}pr^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}pr^\S\,\hat{\gamma^S}\hat{)}\hat{\big)}\,(eq^\A\,\hat{\gamma})\hat{\Big)}\,(u^\S\,\hat{\gamma^S})\hat{\bigg)}\,\hat{(}eq^\S\,\hat{\gamma^S}\hat{)} \\
  & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\S\,\hat{\gamma^S} && :\equiv \\
  & && \hspace{-10em}\hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}\hat{\lambda}\,\hat{p^D}\,\hat{p^S}\hat{.}\,\hat{\refl}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\S\,\hat{\gamma^S}\hat{)}\hat{\big)}\,\hat{(}pr^\S\,\hat{\gamma^S}\hat{)}\\
  & ((\hat{x}\in \hat{A})\ra b)^\S\,\hat{\gamma^S\,f\,t} && :\equiv b^\S\,\hat{\gamma^S}\,\hat{(}\hat{f}\,\hat{t}\hat{)} \\
  & (t\,\hat{u})^\S\,\hat{\gamma^S} && :\equiv \ap\,\hat{(}\hat{\lambda} \hat{f.}\hat{f}\,\hat{u}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)}
\end{alignat*}
\endgroup

>>>>

The present paper is an expanded version of our conference paper \cite{hiit}.
In this version, we extend signatures with paths between type constructors, and
in Section \ref{sec:morphisms} we also compute notions of homomorphisms from
signatures. We also explain a coherence problem in interpreting syntaxes of type
theories, and how it influenced the current paper, in Section
\ref{sec:coherence} and Section \ref{sec:categorical}.

@@

Cartmell's generalized algebraic theories (GATs) \cite{gat} pioneered a
type-theoretic notion of algebraic signature. GATs can be viewed as a variant of
finitary quotient inductive-inductive signatures (QIITs), although GATs also
support equations between sorts (type constructors), which so far have not been
considered in QIIT and HIT literature.

@@

(5) There is a different equality type former, which can be used to express
paths between type constructors. To our knowledge, this has not been considered
so far in HIT literature, although Cartmell considered equations between sorts
in generalized algebraic theories \cite{gat}. For an example, paths between type
constructors allow a compact definition of integers:
\begin{alignat*}{5}
  & \boldsymbol{\cdot},\,\,Int : U,\,\,zero : \underline{Int},\,p : Int =_\U Int
\end{alignat*}
Here, successor and predecessor functions can be recovered from the $p : Int
=_\U Int$ equality, by transporting an $Int$ along $p$ or $p^{-1}$. This
signature could be unfolded to a larger one, by replacing $p : Int =_\U Int$
with an explicit $suc : Int \ra \underline{Int}$ constructor and additional
constructors expressing that $suc$ is an equivalence. The definition of $Int$
with a successor equivalence was previously suggested by Altenkirch, Pinyo
[TODO] and Cavallo [TODO]. Our signature can be viewed as a shorthand for the
unfolded version.

@@

(6) We have a function space where the domain is a type in the external
theory. We distinguish it from (3) by using red brick color in the domain
specification. We specify lists and integers as follows, with integers now
given as quotients of pairs of natural numbers:
\begin{alignat*}{4}
  & \hat{A}\in\Type_{\hat{0}} && \vdash\,\boldsymbol{\cdot},\,\, && List:\U,\,\,\,nil:\underline{List},\,\,\,cons:(\hat{x}\in \hat{A})\ra List\ra\underline{List} \\
  & \hat{\Gamma} && \vdash\,\boldsymbol{\cdot},\,\,&& Int:\U,\,\,\,pair:(\hat{x}\,\hat{y}\in\hat{Nat})\ra\underline{Int},\,\,\, \\
  & && && eq:(\hat{a}\,\hat{b}\,\hat{c}\,\hat{d}\in \hat{Nat})(\hat{p}\in \hat{a}\hat{\mathbin{+}}\hat{d}\eqh_{\hat{Nat}} \hat{b}\hat{\mathbin{+}}\hat{c})\ra \underline{pair\,\hat{a}\,\hat{b}=_{Int}\mathsf{pair}\,\hat{c}\,\hat{d}}, \\
  & && && trunc:(x y : Int)(p\,q : a=_{Int} b)\ra \underline{p=_{x=_{Int} y} q}
\end{alignat*}
In the case of integers, $\hat{\Gamma}$ is
$\hat{Nat}\in\Type_{\hat{0}}\hat{,}\,\blank\hat{\mathbin{+}}\blank\in\hat{Nat}\rah
\hat{Nat}\rah\hat{Nat}$, or alternatively, we could require natural
numbers in the external theory. As another example, propositional
truncation for a type $\hat{A}$ is specified as follows.
\[
\hat{A}\in\Type_{\hat{0}}\vdash\, \boldsymbol{\cdot},\,\,\,tr:\U,\,\,\,emb : (\hat{x}\in \hat{A})\ra \underline{tr},\,\,\,eq:(x\,y:tr)\ra \underline{x=_{tr} y}
\]
The smallness of $\hat{A}$ is required in (5). It is possible to generalize
signatures to arbitrary universe levels, but it is not essential to the current
development. Note that the (5) function space preserves strict positivity, since
in the external theory there is no way to recursively refer to the inductive type
\emph{being defined}. The situation is analogous to the case of $W$-types
\cite{abbot05containers}, where shapes and positions can contain arbitrary types
but they cannot recursively refer to the $W$-type being defined.

(7) $\U$ is also closed under a function space where the domain is an external
type and the codomain is a small source theory type. We overload the application
notation for external parameters, as it is usually clear from context which
application is meant. The rules allow types with infinitary constructors, for
example trees containing elements of $\hat{A}$ at the leaves and branching by
$\hat{B}$ (which could be an infinite type):
\[
\hat{A}\in\Type_{\hat{0}}\commah\hat{B}\in\Type_{\hat{0}}\vdash\,\boldsymbol{\cdot},\,\,\,T:\U,\,\,\,leaf:(\hat{x}\in \hat{A})\ra\underline{T},\,\,\,node:((\hat{x}\in \hat{B})\ra T)\ra\underline{T}
\]
Here, $leaf$ has a function type (5) and $node$ has a function type (3) with a
function type (6) in the domain. More generally, we can define $W$-types
\cite{abbot05containers} as follows. $\hat{S}$ describes the ``shapes'' of the
constructors and $\hat{P}$ the ``positions'' where recursive arguments can
appear.
\[
\hat{S}\in\Type_{\hat{0}}\commah\hat{P}:\hat{S}\rah\Type_{\hat{0}}\vdash\,\boldsymbol{\cdot},\,\,\,W:\U,\,\,\,sup: (\hat{s} \in \hat{S})\ra((\hat{p} \in \hat{P}\,\hat{s})\ra W)\ra \underline{W}
\]
For a more complex infinitary example, see the definition of Cauchy
reals in \cite[Definition 11.3.2]{HoTTbook}. It can be also found as
an example file in our Haskell implementation.

@@

\begingroup
\allowdisplaybreaks
\begin{alignat*}{5}
  & \boldsymbol{\cdot}^\S\,\hat{\gamma}\,\hat{\gamma^D} && :\equiv \hat{\top} \\
  & (\Delta,\,x:A)^\S\,(\hat{\gamma},\hat{t})\,(\hat{\gamma^D},\hat{t^D}) && :\equiv \hat{(}\hat{\gamma^S}\in\Delta^\S\hat{\gamma^2}\hat{)}\timesh A^\S\,\hat{\gamma^S}\,\hat{t}\,\hat{t^D} \\
  & x^\S\,\hat{\gamma^S} && :\equiv x^{\text{th}}\text{ component in } \hat{\gamma^S} \\
  & \U^\S\,\hat{\gamma^S}\,\hat{A}\,\hat{A^D} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah  \hat{A^D}\,\hat{x} \\
  & (\underline{a})^\S\,\hat{\gamma^S}\,\hat{t}\,\hat{t^D} && :\equiv a^\S\,\hat{\gamma^S}\,\hat{t} \eqh \hat{t^D} \\
  & ((x:a)\ra B)^\S\,\hat{\gamma^S}\,\hat{f}\,\hat{f^D} && :\equiv \hat{(}\hat{x}\in a^\A\,\hat{\gamma}\hat{)}\rah  B^\S\,\hat{(}\hat{\gamma,x}\hat{)}\,\hat{(}\hat{\gamma^D,}\,a^\S\,\hat{\gamma^S}\,\hat{x}\hat{)}\,\hat{(}\hat{\gamma^S,}\,\hat{\refl}\hat{)} \\
  & && \hspace{8.4em} \hat{(}\hat{f}\,\hat{x}\hat{)}\,\hat{\big(}\hat{f^D}\,\hat{x}\,\hat{(}a^\S\,\hat{\gamma^S}\,\hat{x}\hat{)}\hat{\big)} \\
  & (t\,u)^\S\,\hat{\gamma^S} && :\equiv \hat{\J}\,\hat{(}t^\S\,\hat{\gamma^S}\,\hat{(} u^\A\,\hat{\gamma}\hat{)}\hat{)}\,\hat{(}u^\S\,\hat{\gamma^S}\hat{)} \\
  & ((\hat{x}\in \hat{A})\ra B)^\S\,\hat{\gamma^S}\,\hat{f}\,\hat{f^D} && :\equiv \hat{(}\hat{x}\in \hat{A}\hat{)}\rah B^\S\,\hat{\gamma^S}\,\hat{(}\hat{f}\,\hat{x}\hat{)}\,\hat{(}\hat{f^D}\,\hat{x}\hat{)} \\
  & (t\,\hat{u})^\S\,\hat{\gamma^S} && :\equiv t^\S\,\hat{\gamma^S}\,\hat{u} \\
  & (t=_a u)^\S\,\hat{\gamma^S}\,\hat{e} && :\equiv \TR\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)}\hat{\big(}\TR\,\hat{(}u^\S\,\hat{\gamma^S}\hat{)}\,\hat{(}\apd\,\hat{(}a^\S\,\hat{\gamma^S}\hat{)}\,\hat{e}\hat{)}\hat{\big)} \\
  & (\refl_t)^\S\,\hat{\gamma^S} && :\equiv \hat{\J}\,\hat{\refl}\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)} \\
  & (\J_{a\,t\,(x.z.p)}\,pr\,_u\,eq)^\S\,\hat{\gamma^S} && :\equiv \\
  & && \hspace{-10em}\hat{\J}\,\hat{\bigg(}\hat{\J}\,\hat{\Big(}\hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}\hat{\lambda}\,\hat{p^D}\,\hat{p^S}\,\hat{pr^D}\,\hat{pr^S}\hat{.}\,\hat{pr^S}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)}\,\\
  & && \hspace{-9em}\hat{(}\hat{\mathsf{uncurry}}\,p^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\S\,\hat{\gamma^S}\hat{)}\,\hat{(}pr^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}pr^\S\,\hat{\gamma^S}\hat{)}\hat{\big)}\,\hat{(}eq^\A\,\hat{\gamma}\hat{)}\hat{\Big)}\,\hat{(}u^\S\,\hat{\gamma^S}\hat{)}\hat{\bigg)}\,\hat{(}eq^\S\,\hat{\gamma^S}\hat{)} \\
  & (\J\beta_{a\,t\,(x.z.p)}\,pr)^\S\,\hat{\gamma^S} && :\equiv \\
  & && \hspace{-10em}\hat{\J}\,\hat{\big(}\hat{\J}\,\hat{(}\hat{\lambda}\,\hat{p^D}\,\hat{p^S}\hat{.}\,\hat{\refl}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\D\,\hat{\gamma^2}\hat{)}\,\hat{(}\hat{\mathsf{uncurry}}\,p^\S\,\hat{\gamma^S}\hat{)}\hat{\big)}\,\hat{(}pr^\S\,\hat{\gamma^S}\hat{)}\\
  & ((\hat{x}\in \hat{A})\ra b)^\S\,\hat{\gamma^S\,f\,t} && :\equiv b^\S\,\hat{\gamma^S}\,\hat{(}\hat{f}\,\hat{t}\hat{)} \\
  & (t\,\hat{u})^\S\,\hat{\gamma^S} && :\equiv \ap\,\hat{(}\hat{\lambda} \hat{f.}\hat{f}\,\hat{u}\hat{)}\,\hat{(}t^\S\,\hat{\gamma^S}\hat{)}
\end{alignat*}
\endgroup
